\MakeUppercase{Layered $H$-Partitions of $k$-Planar Graphs} \thanks{This work was partly funded by NSERC and MRI.}

### 2019-05-16 k-planar.tex
Reading time.....

We prove the following result:
\begin{thm}\thmlabel{k-planar}
  Every $k$-planar graph has a layered $H$-partition of layered-width at most $24k^2 + 36k + 12$ in which $H$ has treewidth at most $k^3/6+k^2/+k/3-1$.
\end{thm}

@@

For each edge $vw\in E(G)$, consider the walk $W_{vw}$ in $G^+$ that corresponds to the sequence of vertices in $G^+$ encountered while walking along the curve $vw$ from $v$ to $w$. We define $A_{vw}=\bigcap_{z\in W_{vw}} P_z$.  Since $V(W_{vw})\subseteq V(G^+)$, $A_{vw}$ is well-defined.  For each $z\in V(W_{vw})$, $K[P_{z}]$ is a vertical path in $K$ that contains $r_K$.   Therefore $A_{vw}$ is a non-empty vertical path in $K$.  For any vertex $v\in V(G)$ define $A_v=\bigcap_{vw\in E(G)} A_{vw}$.  Again, $K[A_v]$ is a non-empty vertical path in $K$ that contains $r_K$.  We define $a(v)$ to be the lower endpoint of $K[A_v]$.  For every node $x\in V(K)$, define the \emph{separator}
\[
   S_x = \{v\in V(G): a(v)=x \} \enspace .
\]
\note{PM}{We might be able to save a factor of 2 here on the layered width, because we only need one endpoint of each edge $vw$.}Clearly $\mathcal{S}=\{S_x:x\in V(K)\}$ is a partition of $V(G)$.  We will use $\mathcal{S}$ as the partition that defines our quotient graph $H=G/\mathcal{S}$ with vertex set $V(H)=V(K)$ and the edge $xy\in E(H)$ if and only if there exists an edge $vw\in E(G)$ with $v\in S_x$ and $w\in S_y$.  First we argue that this partition has small width with respect to the layering $\{V_0,\ldots,V_h\}$ defined by $T$ (recall that this is a layering of $G^+$, not a layering of $G$).

@@

  By \clmref{bag-size}, $\mathcal{S}$ produces a graph $H=G/\mathcal{S}$ of treewidth at most $k^3/6+k^2/+k/3-1$.  By \clmref{width-of-g-plus}, for each $x\in V(H)$ and each integer $i$, $|B_x\cap V_i|\le 24k+12$, so $|B_x\cap V_i'|\le (k+1)(24k + 12) = 24k^2+36k+12$. 
\end{proof}

@@

  Apply \lemref{make-planar} to $G^{++}$ to obtain the set $Z\subseteq V(G^{++})$, the planar graph $G^*$, and the BFS spanning tree $T^*$ rooted at $r^*$.  For each edge $vw\in E(G)$, let $W_{vw}$ be the corresponding walk in $G^+$. The set $Z$ contains vertices in $V(G)$ as well as dummy vertices in $V(G^+-G)$. Create a new set $Z'\subseteq V(G)$ as
  \[
     Z' = \{ v : V(W_{vw})\cap Z\neq\emptyset \} \enspace .
  \]
  \note{PM}{We can certainly save a factor of 2 here because we only need one endpoint of each $vw$ that crosses $Z$ to take part in $Z'$.}
  Since $|Z\cap V_i| \le 2g$, the same argument used in the proof of \lemref{width-of-g-plus} shows that $|Z'\cap V_i|\le 2g\times 4\times (2k+1) = 16gk+8g$ for all $i\in\{1,\ldots,h\}$.

>>>>

We prove the following result:
\begin{thm}\thmlabel{k-planar}
  Every $k$-planar graph has a layered $H$-partition of layered-width at most $24k^2 + 36k + 12$ in which $H$ has treewidth at most $k^3/6+k^2/2+k/3-1$.
\end{thm}

@@

For each edge $vw\in E(G)$, consider the walk $W_{vw}$ in $G^+$ that corresponds to the sequence of vertices in $G^+$ encountered while walking along the curve $vw$ from $v$ to $w$. We define $A_{vw}=\bigcap_{z\in W_{vw}} P_z$.  Since $V(W_{vw})\subseteq V(G^+)$, $A_{vw}$ is well-defined.  For each $z\in V(W_{vw})$, $K[P_{z}]$ is a vertical path in $K$ that contains $r_K$.   Therefore $A_{vw}$ is a non-empty vertical path in $K$.  For any vertex $v\in V(G)$ define $A_v=\bigcap_{vw\in E(G)} A_{vw}$.  Again, $K[A_v]$ is a non-empty vertical path in $K$ that contains $r_K$.  We define $a(v)$ to be the lower endpoint of $K[A_v]$.  For every node $x\in V(K)$, define the \emph{separator}
\[
   S_x = \{v\in V(G): a(v)=x \} \enspace .
\]
Clearly $\mathcal{S}=\{S_x:x\in V(K)\}$ is a partition of $V(G)$.  We will use $\mathcal{S}$ as the partition that defines our quotient graph $H=G/\mathcal{S}$ with vertex set $V(H)=V(K)$ and the edge $xy\in E(H)$ if and only if there exists an edge $vw\in E(G)$ with $v\in S_x$ and $w\in S_y$.  First we argue that this partition has small width with respect to the layering $\{V_0,\ldots,V_h\}$ defined by $T$ (recall that this is a layering of $G^+$, not a layering of $G$).

@@

  By \clmref{bag-size}, $\mathcal{S}$ produces a graph $H=G/\mathcal{S}$ of treewidth at most $k^3/6+k^2/2+k/3-1$.  By \clmref{width-of-g-plus}, for each $x\in V(H)$ and each integer $i$, $|B_x\cap V_i|\le 24k+12$, so $|B_x\cap V_i'|\le (k+1)(24k + 12) = 24k^2+36k+12$. 
\end{proof}

@@

  Apply \lemref{make-planar} to $G^{++}$ to obtain the set $Z\subseteq V(G^{++})$, the planar graph $G^*$, and the BFS spanning tree $T^*$ rooted at $r^*$.  For each edge $vw\in E(G)$, let $W_{vw}$ be the corresponding walk in $G^+$. The set $Z$ contains vertices in $V(G)$ as well as dummy vertices in $V(G^+-G)$. Create a new set $Z'\subseteq V(G)$ as
  \[
     Z' = \{ v : V(W_{vw})\cap Z\neq\emptyset \} \enspace .
  \]
  \note{PM}{We can certainly save a factor of 2 here because we only need one endpoint of each $vw$ that crosses $Z$ to take part in $Z'$.}
  Since $|Z\cap V_i| \le 2g$, the same argument used in the proof of \lemlabel{width-of-g-plus} shows that $|Z'\cap V_i|\le 2g\times 4\times (2k+1) = 16gk+8g$ for all $i\in\{1,\ldots,h\}$.
